---
title: "intro_to_r"
author: "Jessica Kosie"
date: "6/24/2018"
output: html_document
---

# Goals 

By the end of this tutorial, you will know:

+ Basic R usage (using R as a calculator, creating variables, indexing vectors)
+ How to read in and examine data.
+ How to get values out of the rows and columns in your data.
+ What is a pipe and how to use them.
+ How to change and summarize data using tidyr. 


## Basic R Use

R as a calculator

```{r}

# Basic arithmetic
2+3

# Follows order of operations (PEMDAS)
(2^3)+4*(5/3)

```
Assigning values

```{r}

# Create a variable called x, that is assigned the number 8. 
x <- 8 
x = 8

# What value did I assign to x?
x

# Can also assign a range of values
x <- 1:5 #x is now a vector of 1, 2, 3, 4, 5

# note that x is no longer 8, it take on whatever the most recent assignment was

# Can do operations on this vector using function
mean(x) #R does these operations on the set of values
sd(x)
sum(x)

# more complex operations
z <- (x-mean(x)/sd(x))

```
Arrays and counting

```{r}

#let's get the numbers 1 thru 10 by ones 
1:10

#sequence of numbers, 1 thru 10, by 2
seq(1,10,2) #can say "by=" before the 2, but it's not necessary
 
```
> **Exercise.** Create a variable called x that is assigned the number 5. Create a variable called y that is a sequence of numbers ranging from 0 to 25 by 5. Multiply x and y. What happens? 
```{r}

```
Creating and indexing matrices

```{r}
x <- matrix(c(11,12,13,21,22,23), byrow=TRUE,nrow=2) #put into a matrix by row
x1 <- matrix(c(11,12,13,21,22,23), byrow=FALSE,nrow=2) #put into a matrix by column

#indexing matrices (getting the value that's in a particular row/column)

# x[r,c] would give you the element in row r, column c of the matrix

x[2,3] #gives you the element in row 2 column 3 of matrix x (defined above)
x[] #all rows, all columns - could have just typed x
x[1, ] #first row, all columns
x[ ,3] #all rows, third column
x[1,3] #first row, third column

y <- x[1,2] #asign the value in the 1st row 2nd column to a new variable

```
## Reading data into R

First, you'll need to tell R where to look for the data. To do this, you will set your working directory. 

```{r}
setwd("/Users/jessicakosie/Desktop/openscience_tutorial-master/datasets/")

# if you're using a PC:
# setwd("C:/Users/YourName/Desktop/openscience_tutorial-master/datasets/")

# In RStudio: Session > Set Working Directory > Choose Directory

# take a look at what's in your directory

dir()

# Now, let's read in the mental_abacus_data CSV file and save it as an object called ma_data

ma_data <- read.csv("mental_abacus_data.csv", header = T)

```
## Examining the data file

We can simply look at the data frame. We can also get a summary of the data.
```{r}

View(ma_data)
summary(ma_data)

```
> **Exercise.** Read in the pragmatic_scales_data csv file. Save it in an object called ps_data. Get a summary of ps_data. What is the mean age of children in the dataset?
```{r}










```







